From big data to deep insight in developmental science

### 2015-08-26 wiley-big-data-devel/big-data.tex
full draft

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}

@@

\oneauthor{Rick O. Gilmore}
\twoaffiliations{The Pennsylvania State University, The Databrary Project}

\abstract{}

@@

A search on the term ``big data'' yields more than 49 million hits on Google and 13 million hits on Bing.
The results return in less than a second.
The term yields more than 100,000 results on Google Scholar.
A search of the term using Google's ngram viewer (http://books.google.com/ngram) shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
Google's ngram viewer indexes terms in digitized books.
So, whether measured by search engine matches of electronic documents available on the internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
That use has grown substantially over the past several decades.
The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
Most of the tools that enable it have been created in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015; boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.

In this paper, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
Then, I will briefly survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future of the field.

I will argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
Nor, do I find that ``big data'' analyses are all that new in developmental psychology.
Rather, I find that a while the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily interrelatable datasets.
Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual esearcher behavior, publisher priorities, and research funding levels.
Technology may have enabled the ``big data'' era, but the challenges it poses may turn out to be the least important, at least where advancing research in developmental psychology is concerned.

@@

Doug Laney \cite{laney01controlling3v} was the first person to describe the dimensions of data streams that make them especially ``big'': Volume, velocity, and variety.
Data streams that are large in quantity, change rapidly, or consist of many components merit special consideration \cite{ibm_2015}.
Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.

What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
Similarly, the measure of volume matters.
The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{ICPSR}.
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?

The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops? Why does it develop this way?
In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
For social and behavioral scientists, the goal is to describing change in the *behavior* or *beliefs* of individuals or groups of individuals and families.
Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
Similarly, the assessment of biological data -- genes, hormone levels, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status -- inform thinking about how these within-person factors influence changes within individuals' behavior across time.

Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
We will see that some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
Similarly, the aspects of data `velocity' most relevant to developmental researchers are those related to the frequency or spacing of measurements.
Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
`Variety' encompasses the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical/evolutionary and the use of multiple measurements to address the same underlying construct.
To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.

@@

In the next section, I will review the current state of big data in developmental research.
In doing so, I summarize information about the size of datasets along the relevant dimensions we have discussed, the availability of datasets for reuse or reanalysis by others, and the impact of particular datasets on scholarship.

@@

% How big, in what ways are big?
% How open, in what ways open?
% Size of impact on scholarship?
% Domains of interest with big datasets?
% Government vs. individual investigator?

Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities like the U.S. Census Bureau and the U.S. Department of Education.
The collection and dissemination of these datasets is mandated by law, subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
% Mention focus on individuals and families?
The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
% Mention Federal regulations that guide research with human subjects.
Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
But, commercial entities are beginning to collect and store health and fitness related activity (Fitbit, etc.) 
These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities who provide the services.

@@

Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
The National Health and Nutrition Examination Survey (NHANES) \cite{NHANES} began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S. infants, children, and adults. 
The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
Many will be familiar with some of NHANES's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm).
The National Youth Fitness Survey \cite{NYFS} is closely associated with NHANES.
NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
Much of the NHANES and NNYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
A Google Scholar search shows more than 110,000 publications mention the NHANES dataset.

The National Longitudinal Survey of Youth (NLSY) \cite{NLSY} provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52 through the National Longitudinal Survey (NLS).
Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data} to which users must apply for access.
Access to geographic-related variables requires special permission.
The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
A Google Scholar search in mid-2015 shows that more than 20,000 publications cite the NLSY.

The Early Childhood Longitudinal Study (ECLS) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences. 
The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
ECLS consists of both publicly available and restricted data.
A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications (https://www.icpsr.umich.edu/icpsrweb/ICPSR/biblio/series/00191/resources?sortBy=1).

There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years \cite{TIMSS}.
TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commerical partners.
The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
Both studies provide data available for public download.

In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
Launched in 2011, the Japan Environment and Children's Study \cite{JECS} is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study is in its early stages.
The study has produced a handful of publications \cite{JECS_pubs}, and no data sharing plans have been announced.
Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health & Development \cite{NSHD_UK} and the National Child Development Study (NCDS) \cite{NCDS_UK}.
NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
Measures include genetic, health, education, social, and economic factors.
The study is funded by the British government and data are available by registration from the U.K. Data Service \cite{UK_cohort_data}.
The British Cohort Studies have generated more than 3,500 publications to-date \cite{UK_cohort_citations}.
Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset \cite{PISA}, which generates more than 5,000 search hits.

Not all large-scale birth cohort studies succeed.
A notable failure in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel. 
Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
Those data are slated for release in a data archive sometime in 2015.

While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
We turn next to datasets whose collection and management is under the control of individual, usually academic investigators.

\subsection{Academic Investigator Initiated and Managed Datasets}

Investigator-managed big data projects cover a wide range of topics and and even wider range of research methodologies.
In several cases, the studies involve partnerships between government agencies and academic investigators.
The Panel Study on Income Dynamics (PSID) \cite{PSID} is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
Much of data is available publicly to registered users and agree to the terms of use conditions \cite{PSID_term} that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement \cite{PSID_restricted_use} between the investigator and the data owner, the University of Michigan.

The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} began as a survey study designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year. 
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
The Ad Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.

The NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, and approval is granted on a case-by-case basis.
Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.

The Family Life Project (FLP) studied the early development children living in rural, largely poor communities in North Carolina and Pennsylvania. 
A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
These features make the FLP large in volume, velocity, and variety.
Data from Phase I of the FLP are archived at ICPSR \cite{Family_Life_data}, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
The FPL site lists about 50 publications through 2013.

@@

New initiatives like DataWallet (https://www.datawallet.io/) are on the horizon that may empower individuals to share their data for a fee.

>>>>

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[hyphens]{url}
\usepackage{hyperref}

\hypersetup{breaklinks=true}

@@

\author{Rick O. Gilmore}
\affiliation{{The Pennsylvania State University}, {The Databrary Project}}

%\abstract{}

@@

A search on the term ``big data'' yields more than 49 million hits on Google, more than 100,000 results on Google Scholar, and 13 million hits on Bing.
The results return in less than a second.
A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
So, whether measured by search engine matches of electronic documents available on the Internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
That use has grown substantially over the past several decades.
The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
Most of the tools that enable it have been created in the last 20 years.
``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.

In this review, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
Then, I survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
I close by discussing a range of issues that the emergence of ``big data'' approaches to developmental science pose for the future of the field.

I argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
Nor are ``big data'' analyses all that new in developmental psychology.
Rather, while some the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
Technology may accelerate the ``big data'' era, but the challenges it poses may turn out to be the least important where advancing research in developmental psychology is concerned.

@@

According to Laney \citeyear{laney01controlling3v} the volume, velocity, and variety of data streams make data big.
Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
Similarly, the measure of volume matters.
The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the large number of files in the repository totaled about 10 TB of digital storage.

Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops? Why does it develop this way?
In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
For social and behavioral scientists, the goal is to describe change in the \emph{behavior} or \emph{beliefs} of individuals or groups of individuals and families.
Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
Similarly, the assessment of biological data -- genes, hormone levels, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status -- inform thinking about how these within-person factors influence changes within individuals' behavior across time.

Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
Some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
Similarly, the aspects of data `velocity' most relevant to developmental researchers are those related to the frequency or spacing of measurements.
Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
`Variety' encompasses the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical/evolutionary and the use of multiple measurements to address the same underlying construct.
To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.

@@

@@

Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities like the U.S. Census Bureau and the U.S. Department of Education.
The collection and dissemination of these datasets is mandated by law, subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities that provide the services.

@@

Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S. infants, children, and adults. 
The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
Many will be familiar with some of NHANES's products such as standardized child growth charts, and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}).
NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
Much of the NHANES and NYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
A Google Scholar search in mid-2015 shows more than 110,000 publications mention the NHANES dataset.

The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52.
Most NLSY and NLS data are available free of charge to by means of a web-based data portal to which users must apply for access.
Access to geographic-related variables requires special permission.
The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
A Google Scholar search in mid-2015 shows that more than 20,000 publications cite the NLSY.

The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences. 
The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
ECLS consists of both publicly available and restricted data.
A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.

There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}).
TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commercial partners.
The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
Both studies provide data available for public download.

In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
Launched in 2011, the Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study has produced a handful of publications, and no data sharing plans have been announced.
Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
Measures include genetic, health, education, social, and economic factors.
The study is funded by the British government and data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
The British Cohort Studies have generated more than 3,500 publications to-date.
Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.

Not all large-scale birth cohort studies succeed, of course.
A notable failure in the U.S. is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel. 
Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
Those data are slated for release in a data archive sometime in 2015.

While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
We turn next to datasets whose collection and management is under the control of individual, usually academic investigators.

\subsection{Public/Private Partnerships and Investigator-driven Project}

Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
Others involve large-scale data collections that are initiated and managed by individual investigators.
These projects are diverse in focus and methodology, including population-based studies with a health focus, to behavior genetics, brain imaging, cognition and temperament, and education and employment.

\subsubsection{Population-Based Studies}

A population-based example is the Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
As of mid-2015, PSID had generated more than 3,900 publications. 
Much of data is available publicly to registered users and agree to the terms of use conditions that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.

The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) is another population-based public/private partnership.
SECCYD prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometric and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
The SECCYD has generated more than 214 publications as of mid-2015.

The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) represents a third example.
Ad Health began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
The study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year.
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
The Ad Health data are archived at ICSPR and are available for public use.
As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.

Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) took a population-based approach to studying the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania. 
A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
The FLP site lists about 50 publications through 2013.

@@
